Probabilistic Inference of NYC Congestion from Taxi Data

### 2017-12-10 reports/paper/paper.tex
Paper edits second half.

In addition to adding columns, we filtered the original dataset. We only consider trips
whose start- and end-locations that falls inside a rectangle that bounds Manhattan. We also removed trips with duration less than two minutes or more than two hours.

We performed a principal component analysis on all de-meaned and
standardized geographic coordinates (both start
and end locations) from the filtered data. We thus transform the coordinates into the first and second principal components. This allowed us to compute the rotational matrix so that Manhattan lies approximately orthogonal to the $x$-axis. For the discrete
grid models, these coordinates were further discretized into a $20 \times 70$
grid. Each node in the grid approximately corresponds to a $400\text{ft} \times 400\text{ft}$ area. We chose the grid size as a compromise between computational tractability and model expressiveness.

We saved the dataset as a whole, and we also partitioned into $(\text{day of week},
\text{hour})$ groups, using the one-hot encoded vectors. Some models were
trained on the entire dataset (e.g.\ Neural Network), and some were trained
individually on the partioned datasets (e.g.\ Structural Models).

@@

We calculated three baselines. The first was a linear regression predicting
trip duration from trip distance (calculated using the Pythagorean Theorem and
the start and end coordinates), and trained on a partitioned data set.

@@

The third baseline was constructed by querying the Google Maps Distance Matrix
API.\footnote{\url{https://developers.google.com/maps/documentation/distance-matrix/}}
While there was no need to train this baseline, there was a rate limit
constraint on how many test data points it could be tested on. We assess its
performance on approximately 2,500 data points in a representative partioned test dataset. Moreover, Google Maps API does not allow predicting travel time in the past, and thus for each trip, we use a random Tuesday 8:00 AM in January 2018 as a proxy. From our tests, the predictions of Google Maps API are not significantly different, whether the user submits a start time or not. Thus, we believe that this is a reasonable approximation given limitations to the data.

@@

Training implementations varied from model to model. In training the neural
network baseline, SGD was used with a learning rate of $10^{-7}$ run for 12
epochs on the full data set. This took between 30 minutes and an hour. The
uniform paths structural model was similarly trained with SGD, with a learning
rate of $10^{-4}$.

The uniform model was only trained on a subset of the
dataset corresponding to all trips on a given day of the week and on a given
hour. Convergence appeared after roughly two epochs, and the model was run for 5
epochs taking just under ten minutes.

The softmax regression model was trained
using the Adam optimizer \cite{kingma2014adam} with learning rate $5\times 10^{-3}$ for various epochs for different data-subsets,\footnote{The epochs are calculated so that the total numbers of trips trained by the model are approximately constant across the subsets of data. This means that we trained for 131 epochs for 4:00 AM but only 6 epochs for 8:00 AM, since the two data-subsets have drastically different number of trips. Nonetheless, as Figure~\ref{fig:softmaxloss} shows, we reached convergence in each dataset we trained on.}
taking approximately thirty minutes for each data-subset. Additionally, the softmax regression model take hyperparameters $L$ (sample size in each stochastic draw) and $\sigma^2$ (variance of trip time). We chose $L = 20$ and $\sigma^2 = 100$. These values are again chosen to strike a bargain between realism and computation feasibility. The model appears to converge by the second epoch.

@@

We train our model and test the model on an out-of-sample test-set using the methods discussed in Section~\ref{sub:predict}. As a simple check of the training process, we plot in Figure~\ref{fig:softmaxloss} the loss obtained when fitting the softmax regression model. We see that the model converges relatively fast.\footnote{The full fitting takes approximately 30 minutes per dataset (2.9 GHz Intel Core i5, 8GB RAM), but the model appears to converge in only 20\% of that time.} The convergence plot for the uniform model is extremely similar and thus omitted. Table~\ref{tab:mainres} contains the main prediction results of our model relative to some benchmark predictions
such as the Google Maps Distance Matrix API, neural network, and linear regression. We note that, in terms of predictive metrics, our models were able to outperform Google Maps and linear regression, but were not able to outperform the neural network, even though the difference is not very large. Despite the disappointing outcome, the advantage of our framework is that the parameters have structural interpretations, whereas the neural network is a black box.

@@

Table~\ref{tab:mainres} is encouraging. Our models successfully leverage the power of geography and picks up more signal in the data than distance of the trip alone, as demonstrated by the significantly improved performance over linear regression. Perhaps more strikingly, our model slightly outperforms the predictions generated by Google Maps, which means that the predictive accuracy of the model is on par with the standard of a widely used consumer product. Even though we were not able to outperform the neural network, we reiterate that the main advantage of our approach is to have interpretable parameters, which are visualized in Figure~\ref{fig:unifmap}. We turn to the figure in the next section.
% \begin{itemize}
% \item What conclusions can you draw from the results section?
% \item Is there further analysis you can do into the results of the system? Here is a good place to include visualizations, graphs, qualitative analysis of your results.
% \item  What questions remain open? What did you think might work, but did not?
% \end{itemize}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{figs/unifmaps.pdf}
\includegraphics[width=0.5\textwidth]{figs/softmaxmaps.pdf}
\caption{Learned weights, various times Tuesday. Top row: uniform model.
Bottom row: softmax model.}
\label{fig:unifmap}
\end{figure}

@@

Figure~\ref{fig:unifmap} shows that the parameter values for the uniform and softmax regression models behave similarly, with similar areas of high and low congestion. The patterns seem to largely agree with conventional wisdom, both in spatial and temporal dimensions. We identify that 4:00 AM is indeed less congested than rush hour, that Midtown Manhattan seems to be particularly congested during the day, and that the congested areas have an average speed of 4--6 miles per hour. The visualization also confirms some less-than-well-known features; the areas that demonstrate congestion at 4:00 AM largely confirms with anecdotal evidence.\footnote{According to \textit{Business Insider} (\url{http://read.bi/164vSK1}), East Village and surrounding areas rank highly for nightlife. %#GETLIT
} As the models seem to produce largely similar output, we conclude that they are fairly robust.

@@

While many of our results make sense in the context of the models that produce
them, some puzzles---both qualitative and quantitative---remain. For instance,
visual inspection of the grid of weights that the softmax model converges to
(see bottom row of Figure \ref{fig:unifmap}) reveals a pattern that resembles a
checkerboard. This behavior is most visible in the 4:00 AM weights, but also
appears in certain areas of the other grids. We suspect this has to do with the
superimposition of paths that we train on, but the result largely remains an
unexpected surprise.

Furthermore, we are also intrigued by the bias our models introduce: both the
uniform and softmax structural models on average underestimate trip duration
(see Table \ref{tab:mainres}). Since we optimize an approximate bound, it
is believable that this procedure introduced bias. Nevertheless, we are
missing a good intuition as to why we systematically underestimate and whether
this is a feature or a bug.

@@

In thinking of ways to develop our work further, a couple of directions come to
mind. Foremost among them is taking a closer look at which of the softmax and
uniform models best captures reality. They perform very similarly on our
test data, but their weights correspond to structurally different maps of
Manhattan. Future work could help determine which of these depictions best
tracks actual traffic conditions.

@@

>>>>

In addition to adding columns, we filtered the original dataset. We only
considered trips
whose start- and end-locations fall inside a rectangle that bounds Manhattan. We also removed trips with duration less than two minutes or more than two hours.

We performed a principal component analysis on all de-meaned and
standardized geographic coordinates (both start-
and end-locations) from the filtered data. We transformed the coordinates into
the first and second principal components. This allowed us to compute the
rotational matrix so that the major axis of Manhattan lies approximately orthogonal to the $x$-axis. For the discrete
grid models, these coordinates were further discretized into a $20 \times 70$
grid. Each node in the grid approximately corresponds to a $400\text{ft} \times 400\text{ft}$ area. We chose the grid size as a compromise between computational tractability and model expressiveness.

We saved the dataset as a whole, and we also partitioned into $(\text{day of week},
\text{hour})$ groups, using the one-hot encoded vectors. Some models were
trained on the entire dataset (e.g.\ Neural Network), and some were trained
individually on the partioned datasets (e.g.\ our structural models).

@@

We calculated three baselines. The first was a linear regression predicting
trip duration from Euclidian distance, and trained on a partitioned data set.

@@

The third baseline was constructed by querying the Google Maps Distance Matrix
API.\footnote{\url{https://developers.google.com/maps/documentation/distance-matrix/}}
While there was no need to train this baseline, there was a rate limit
constraint on how many test data points it could be tested on. We assessed its
performance on approximately 2,500 data points in a representative partioned
test dataset. Moreover, Google Maps API does not allow predicting travel time in
the past, and thus for each trip, we used a random Tuesday 8:00 AM in January 2018 as a proxy. From our tests, the predictions of Google Maps API are not significantly different, whether the user submits a start time or not. Thus, we believe that this is a reasonable approximation given limitations to the data.

@@

Training implementations varied from model to model. In training the neural
network baseline, we used SGD with a learning rate of $10^{-7}$ run for twelve 
epochs on the full data set. This took between 30 minutes and an hour. The
uniform paths structural model was similarly trained with SGD, with a learning
rate of $10^{-4}$.

The uniform model was only trained on a subset of the
dataset corresponding to all trips on a given day of the week and on a given
hour. Convergence appeared after roughly two epochs, and the model was run for
five epochs taking just under ten minutes.

The softmax regression model was trained
using the Adam optimizer \cite{kingma2014adam} with learning rate $5\times 10^{-3}$ for various epochs for different data-subsets,\footnote{The epochs are calculated so that the total numbers of trips trained by the model are approximately constant across the subsets of data. This means that we trained for 131 epochs for 4:00 AM but only 6 epochs for 8:00 AM, since the two data-subsets have drastically different number of trips. Nonetheless, as Figure~\ref{fig:softmaxloss} shows, we reached convergence in each dataset we trained on.}
taking approximately thirty minutes for each data-subset. Additionally, the
softmax regression model takes hyperparameters $L$ (sample size in each
stochastic draw) and $\sigma^2$ (variance of trip time). We chose $L = 20$ and
$\sigma^2 = 100$. These were again chosen to strike a bargain between realism
and computation feasibility. The model appeared to converge by the second epoch.

@@

We train our model and test the model on an out-of-sample test-set using the
methods discussed in Section~\ref{sub:predict}. As a simple check of the
training process, we plot in Figure~\ref{fig:softmaxloss} the loss obtained when
fitting the softmax regression model. We see that the model converges relatively
fast.\footnote{The full fitting takes approximately 30 minutes per dataset (2.9
GHz Intel Core i5, 8GB RAM), but the model appears to converge in only one-fifth
of this time.} The convergence plot for the uniform model is extremely similar and thus omitted. Table~\ref{tab:mainres} contains the main prediction results of our model relative to some benchmark predictions
such as the Google Maps Distance Matrix API, neural network, and linear regression. We note that, in terms of predictive metrics, our models were able to outperform Google Maps and linear regression, but were not able to outperform the neural network, even though the difference is not very large.

@@

Table~\ref{tab:mainres} is encouraging. Our models successfully leverage the
power of geography and pick up more signal in the data than distance of the trip
alone, as demonstrated by the significantly improved performance over linear
regression. Perhaps more strikingly, our model slightly outperforms the
predictions generated by Google Maps, which means that the predictive accuracy
of the model is on par with the standard of a widely used consumer product. Even
though we were not able to outperform the neural network, we reiterate that the
main advantage of our approach is to have interpretable parameters, which are
visualized in Figure~\ref{fig:unifmap}. We turn to the figure now.
% \begin{itemize}
% \item What conclusions can you draw from the results section?
% \item Is there further analysis you can do into the results of the system? Here is a good place to include visualizations, graphs, qualitative analysis of your results.
% \item  What questions remain open? What did you think might work, but did not?
% \end{itemize}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{figs/unifmaps.pdf}
\includegraphics[width=0.5\textwidth]{figs/softmaxmaps.pdf}
\caption{Learned weights, various times Tuesday. Top row: uniform model.
Bottom row: softmax model. \emph{In the grids, the southern border of Central Park lies
approximately three-sevenths of the way up}}
\label{fig:unifmap}
\end{figure}

@@

Figure~\ref{fig:unifmap} shows that the parameter values for the uniform and softmax regression models behave similarly, with similar areas of high and low congestion. The patterns seem to largely agree with conventional wisdom, both in spatial and temporal dimensions. We identify that 4:00 AM is indeed less congested than rush hour, that Midtown Manhattan seems to be particularly congested during the day, and that the congested areas have an average speed of 4--6 miles per hour. The visualization also confirms some less-than-well-known features; the areas that demonstrate congestion at 4:00 AM largely confirms anecdotal evidence.\footnote{According to \textit{Business Insider} (\url{http://read.bi/164vSK1}), East Village and surrounding areas rank highly for nightlife. %#GETLIT
} As the models seem to produce largely similar output, we conclude that they are fairly robust.

@@

While many of our results make sense in the context of the models that produce
them, some puzzles---both qualitative and quantitative---remain. For instance,
visual inspection of the grid of weights that the softmax model converges to
(see bottom row of Figure \ref{fig:unifmap}) reveals a pattern that resembles a
checkerboard. This behavior is most visible in the 4:00 AM weights, but also
appears in certain areas of the other grids. We suspect this has to do with the
superimposition of paths that we train on, but the result largely remains a
surprise.

Furthermore, we are also intrigued by the bias our models introduce: both the
uniform and softmax structural models on average underestimate trip duration
(see Table \ref{tab:mainres}). Since we optimize an approximate bound, it
is believable that this procedure introduced bias. Nevertheless, we are
missing a good intuition as to why we systematically underestimate and whether
this is a feature or a bug. It is also interesting to note that the Google Maps
predictions exhibit a bias in the opposite direction.

@@

We now discuss a few directions for future work. Foremost among them is
taking a closer look at which of the softmax and
uniform models best captures reality. They perform very similarly on our
test data, but their weights correspond to structurally different maps of
Manhattan. Future work could help determine which of these depictions best
tracks actual traffic conditions.

@@

Finally, we note that our models condition only on hour and day of the week when
making predictions. However, our framework can incorporate a wider range of
relevant information. For instance, if weather data were accessible, the effect
of different weather conditions on road congestion could be measured. Similarly,
abnormal traffic conditions on holidays could be identified.
